Insecticide resistance evolution with mixtures and sequences : a model-based graphical introduction and summary

### 2017-04-12 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.tex
paper writing

---
title: "Insecticide resistance evolution with mixtures and sequences : a model-based graphical introduction and summary"
output: 
  pdf_document:
    #citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex

#output: word_document
author: "Andy South and Ian Hastings"
date: "`r Sys.Date()`"
bibliography: library.bib
csl: biomed-central.csl
fontsize: 11pt
spacing: double
# below worked in test_linenumbers.Rmd
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
---

```{r, eval=FALSE, include=FALSE}
#!!!! paper protocol
#!!!! might be easier to put text and figs into separate Rmd files after draft stage
#!!!! drafts knit to pdf with pdf_document in YAML and this chunk eval=FALSE
#!!!! submission MS, knit to word_document, YAML, this chunk eval=FALSE, 
#!!!!  in WORD: delete figs, add fig numbers to fig legends, set double space, add line numbers 
#!!!! submission figs, set this chunk to eval=TRUE

# options to create final publication quality figures
# causes the document to fail, but the figs do get stored in pap2figs/
library(knitr)
opts_chunk$set(dev="tiff",
               dev.args=list(compression="lzw"),
               dpi=300,
               cache=FALSE,
               fig.path='pap2figs/')
```

```{r,echo=FALSE,results='hide',warning=FALSE}
  library(resistance)
  library(ggplot2)
```

### version 12

###alternative titles:
###Evolution of insecticide resistance with insecticide mixtures and sequences : an accessible model to promote understanding.
###An accessible model to promote understanding of the evolution of insecticide resistance : insecticide effectiveness predicted to be key to the relative performance of mixtures.
###A tutorial, primer ...
###Insecticide effectiveness predicted to be most important factor determining whether insecticide mixtures can slow evolution of resistance.

@@

## Abstract

_must be 350 words, currently 351_

##### Background
Insecticide resistance threatens the control of vectors and particularly malaria. New insecticides are being developed to address this. Potential strategies for using new insecticides include mixtures, sequences and rotations and there are recommendations for how best to implement these to manage resistance. There is, however, limited recent, accessible modelling work addressing the evolution of resistance under different operational strategies. Previous work concludes that preferred strategies will be situation specific. There is potential to improve the level of mechanistic understanding within the operational community of how insecticide resistance can be expected to evolve in response to different strategies.

@@

##### Results
We use the model to demonstrate the evolution of resistance under different inputs and how this fits with intuitive reasoning about the selection pressure exerted by insecticides. We show that an insecticide used in a mixture, relative to alone, always prompts slower evolution of resistance, but resistance to the two insecticides may evolve more slowly when used in sequence. We show that the ability of insecticides to kill susceptible mosquitoes (effectiveness), has the most influence on whether resistance to two insecticides is likely to arise faster in a mixture or sequence.

##### Conclusions
Our model makes more open the process of insecticide resistance evolution and how it is likely to respond to insecticide use. We provide a simple user-interface allowing further exploration (https://andysouth.shinyapps.io/MixSeqResist1). These tools can contribute to operational decisions in insecticide resistance management.

## Keywords
insecticide resistance; public health; mosquitoes; vector-borne diseases; infectious diseases; malaria; population genetics

## Background

Insecticide resistance is a problem for malaria [@WHO2012][@Ranson2016][@Hemingway2016] other vector borne diseases [@IRAC2011] and agriculture [@FAO2012]. Malaria alone still results in hundreds of thousands of deaths per year. Recent malaria control efforts have centred on treated bed nets and indoor residual spraying, both reliant on insecticides. Treated nets were recently estimated to contribute 68% and indoor residual spraying 13% to averting more than 500 million falciparum malaria cases between 2000 and 2015 [@Bhatt2015]. A recent malaria transmission model [@Churcher2016] predicts that even low pyrethroid resistance is likely to increase malaria incidence in Africa by reducing the performance of bed nets.

The WHO produced a Global Plan for Insecticide Resistance Management in malaria vectors (GPIRM)[@WHO2012] which includes recommendations on operational strategies for managing resistance including the use of insecticide mixtures when they become available.  Efforts are under way to develop new insecticides that will be effective in the light of existing resistance and allow additional options within insecticide resistance management. The Innovative Vector Control Consortium (IVCC) was set up in 2005 to develop new vector control tools and particularly new insecticides to address insecticide resistance in malaria transmitting mosquitoes [@Hemingway2006][@IVCC2016]. Three new insecticides are now in development [@IVCC2016] and likely to be available around 2020 [@Ranson2016]. It is important that decisions about how best to use the new insecticides to delay the onset of resistance are made before insecticides are released [@Hemingway2016].

Modelling studies have investigated the evolution of insecticide resistance in insecticide mixtures including in a public health context e.g. [@Curtis1985][@Mani1985][@Roush1989] but much of the work was done more than 20 years ago and there remained some confusion about the results [@Levick2017]. In an earlier paper [@Levick2017] we described the technical details of a flexible model used to investigate the relative benefits of mixtures and sequences. Here we provide an accessible summary of the model and use selected parameter values to describe mechanistically how the evolution of resistance is influenced by different inputs. This mechanistic understanding can contribute to the debate on the relative merits of different insecticide strategies extending existing frameworks [@Roush1989][@IRAC2011][@FAO2012][@WHO2012].

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. It is likely that genes conferring resistance are present in populations even prior to exposure to novel insecticides thus leading to the potential for selection [@Liu2015].

## Methods

@@

**Table 1. Parameters inflencing the development of insecticide resistance**

Parameter  | Description
---------------------- | ----------------------------------------------------
1. Effectiveness | proportion of susceptible (SS) insects killed by exposure to insecticide
2. Exposure | proportion of insects exposed to insecticide
3. Resistance restoration  | ability of resistance (RR) to restore fitness of insects exposed to insecticide
4. Dominance of restoration | sets fitness of heterozygous (SR) insects between that of SS & RR in presence of insecticide
5. Frequency  | frequency of resistance alleles within the population
6. Cost of resistance  |  fitness of resistant (RR) insects in absence of insecticide
7. Dominance of cost  | sets fitness of heterozygous (SR) insects between that of SS & RR in absence of insecticide

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 1. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1 (equivalent to the unexposed SS), a value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. The 'selection coefficient' of resistance, that may be seen referred to elsewhere [@Curtis1985], can be calculated by multiplying resistance restoration by insecticide effectiveness. Cost of resistance determines the fitness of RR in the portion of the population not exposed to the insecticide.

@@

The model outputs the change in resistance allele frequencies over time measured in generations. We extract the number of generations at which a resistance allele frequency of 50% is reached and term this 'time-to-resistance'. Using a 50% threshold is consistent with previous modelling analyses e.g. [@Curtis1985][@Birget2015]. Preliminary investigations showed that using other potential thresholds of 25 or 75% gave very similar results.

@@

1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the frequency of resistance alleles reaches a threshold of 50%
1. two insecticides used in a mixture (concurrently)

The model is implemented in R [@RCoreTeam2016], the code is hosted on Github [@South2017] and we provide an on-line user interface [@South2017a] (Fig 2) enabling the reader to change inputs and run the model themselves.

## Results

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 3 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 4A) through the even simpler mechanism that less change is required to reach the threshold.

@@

### Two insecticides
When two insecticides are used we can compare the relative performance of a mixture with sequential use. Each panel in Figures 5-9 compares mixture and sequence for a single combination of inputs. To help understand the mechanisms influencing this relative performance we start with a single base scenario and demonstrate changing inputs individually. In this base scenario we set effectiveness, exposure, resistance restoration, and dominance at 0.5, the starting frequencies of resistance to 0.01 and cost of resistance to 0. Resistance arises slower for sequential than mixture for this base scenario (Fig 5A). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer for both to reach the resistance threshold (Fig 5A). Thus the time-to-resistance for the mixture divided by the sequence is less than 1 at 0.8.

#### Insecticide exposure and effectiveness

@@

The results is that a more effective insecticide increases time-to-resistance when used in a mixture (compare the solid lines in Fig 5A & B). This is opposite to what happens when used in sequence (compare the red dashed line in Fig 5A & B) and in isolation (Fig 2A), when more effective insecticides shortened times-to-resistance.

@@

#### Resistance restoration and dominance
Increasing resistance restoration or its dominance decreases time-to-resistance for both sequences and mixtures (Fig 7) in the same way that it did for sole use (Fig 3). The result is that increasing either or both of dominance and resistance restoration (Figs 7B,C,D) doesn't change the relative time-to-resistance for mixtures and sequences from that in the base scenario (Fig 7A) where it takes resistance longer to develop for the sequence. So, unlike effectiveness and exposure, the levels of resistance restoration and dominance do not effect whether a mixture or a sequence are likely to be favoured.

#### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 8C to 8A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 8B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 8D).

#### Costs of resistance
Increasing the cost of resistance to one insecticide in a mixture leads resistance to that insecticide to evolve more slowly (compare the solid red line to the solid blue line in Figure 9C, the red line is for the resistance with the higher cost). The resistance with the higher cost also increases more slowly in the mixture relative to when it is used in sequence (compare the solid red line to the dotted red line in Fig 9C). The greater effect of cost on mixtures means the advantage of sequences over mixtures in Fig 9A (and Fig 5A) is removed in Fig 9C. With costs set to 0 time-to-resistance for mixture divided by sequence is 0.8 in Fig 9A and 1 in Fig 9C. At higher insecticide effectiveness the benefit of mixture over sequence is improved as seen by comparing Fig 9D to 9B.

@@

#### Linkage disequilibrium
It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through a process termed linkage disequilibrium. Figure 8 C&D show how our model predicts that linkage disequilibrium does not lead to more rapid evolution of resistance to the newer insecticide. In Figures 8 C&D the newer insecticide with lower starting resistance (in red) leads to faster resistance when used alone (red dotted) than when used in a mixture (red solid).

## Discussion

@@

These mechanisms, explaining the response of resistance to single insecticide use under different input values, are summarised in Table 2. Referring to figure 1 can help explain these. Exposure and effectiveness act by decreasing the fitness of all genotypes, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR for those not exposed.

#### Two insecticides
Using two insecticides either in sequence or in mixtures differs from the use of a single insecticide and, where it differs can again help us to understand the dynamics of resistance evolution. Our base scenario (Fig 5A) shows curves for resistance to both insecticides in the mixture that are shallower than those for sequential use. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 5A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture, so the sequence is still favoured because 2x60 is greater than 90. Incraesing the effectiveness of on insecticide to Fig 5B decreases the added times for the sequence to 90 and increases those for the mixture to 115 thus changing the order and favouring the mixture.

The mechanism for this change can be sen by comparing the shape of the curves in Figs 5A & B. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide has become ineffective due to resistance it stops killing individuals that are resistant to the second insecticide, loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B.

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide. The more effective insecticide prompts both a) higher selection pressure for it's own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide.

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 5A to 40 generations in Fig 5C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 5B & 5C). This difference, that exposure reduces times-to-resistance for a mixture and effectiveness increases it, was unexpected for us.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Insecticide resistance evolution with mixtures and sequences : a model-based graphical introduction and summary},
            pdfauthor={Andy South and Ian Hastings},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Insecticide resistance evolution with mixtures and sequences : a
model-based graphical introduction and summary}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Andy South and Ian Hastings}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{2017-04-11}

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

\begin{document}
\maketitle

\subsubsection{version 12}\label{version-12}

\subsubsection{alternative titles:}\label{alternative-titles}

\subsubsection{Evolution of insecticide resistance with insecticide
mixtures and sequences : an accessible model to promote
understanding.}\label{evolution-of-insecticide-resistance-with-insecticide-mixtures-and-sequences-an-accessible-model-to-promote-understanding.}

\subsubsection{An accessible model to promote understanding of the
evolution of insecticide resistance : insecticide effectiveness
predicted to be key to the relative performance of
mixtures.}\label{an-accessible-model-to-promote-understanding-of-the-evolution-of-insecticide-resistance-insecticide-effectiveness-predicted-to-be-key-to-the-relative-performance-of-mixtures.}

\subsubsection{A tutorial, primer \ldots{}}\label{a-tutorial-primer}

\subsubsection{Insecticide effectiveness predicted to be most important
factor determining whether insecticide mixtures can slow evolution of
resistance.}\label{insecticide-effectiveness-predicted-to-be-most-important-factor-determining-whether-insecticide-mixtures-can-slow-evolution-of-resistance.}

@@

\subsection{Abstract}\label{abstract}

\emph{must be 350 words, currently 351}

\subparagraph{Background}\label{background}

Insecticide resistance threatens the control of vectors and particularly
malaria. New insecticides are being developed to address this. Potential
strategies for using new insecticides include mixtures, sequences and
rotations and there are recommendations for how best to implement these
to manage resistance. There is, however, limited recent, accessible
modelling work addressing the evolution of resistance under different
operational strategies. Previous work concludes that preferred
strategies will be situation specific. There is potential to improve the
level of mechanistic understanding within the operational community of
how insecticide resistance can be expected to evolve in response to
different strategies.

@@

\subparagraph{Results}\label{results}

We use the model to demonstrate the evolution of resistance under
different inputs and how this fits with intuitive reasoning about the
selection pressure exerted by insecticides. We show that an insecticide
used in a mixture, relative to alone, always prompts slower evolution of
resistance, but resistance to the two insecticides may evolve more
slowly when used in sequence. We show that the ability of insecticides
to kill susceptible mosquitoes (effectiveness), has the most influence
on whether resistance to two insecticides is likely to arise faster in a
mixture or sequence.

\subparagraph{Conclusions}\label{conclusions}

Our model makes more open the process of insecticide resistance
evolution and how it is likely to respond to insecticide use. We provide
a simple user-interface allowing further exploration
(\url{https://andysouth.shinyapps.io/MixSeqResist1}). These tools can
contribute to operational decisions in insecticide resistance
management.

\subsection{Keywords}\label{keywords}

insecticide resistance; public health; mosquitoes; vector-borne
diseases; infectious diseases; malaria; population genetics

\subsection{Background}\label{background-1}

Insecticide resistance is a problem for malaria {[}1{]}{[}2{]}{[}3{]}
other vector borne diseases {[}4{]} and agriculture {[}5{]}. Malaria
alone still results in hundreds of thousands of deaths per year. Recent
malaria control efforts have centred on treated bed nets and indoor
residual spraying, both reliant on insecticides. Treated nets were
recently estimated to contribute 68\% and indoor residual spraying 13\%
to averting more than 500 million falciparum malaria cases between 2000
and 2015 {[}6{]}. A recent malaria transmission model {[}7{]} predicts
that even low pyrethroid resistance is likely to increase malaria
incidence in Africa by reducing the performance of bed nets.

The WHO produced a Global Plan for Insecticide Resistance Management in
malaria vectors (GPIRM){[}1{]} which includes recommendations on
operational strategies for managing resistance including the use of
insecticide mixtures when they become available. Efforts are under way
to develop new insecticides that will be effective in the light of
existing resistance and allow additional options within insecticide
resistance management. The Innovative Vector Control Consortium (IVCC)
was set up in 2005 to develop new vector control tools and particularly
new insecticides to address insecticide resistance in malaria
transmitting mosquitoes {[}8{]}{[}9{]}. Three new insecticides are now
in development {[}9{]} and likely to be available around 2020 {[}2{]}.
It is important that decisions about how best to use the new
insecticides to delay the onset of resistance are made before
insecticides are released {[}3{]}.

Modelling studies have investigated the evolution of insecticide
resistance in insecticide mixtures including in a public health context
e.g. {[}10{]}{[}11{]}{[}12{]} but much of the work was done more than 20
years ago and there remained some confusion about the results {[}13{]}.
In an earlier paper {[}13{]} we described the technical details of a
flexible model used to investigate the relative benefits of mixtures and
sequences. Here we provide an accessible summary of the model and use
selected parameter values to describe mechanistically how the evolution
of resistance is influenced by different inputs. This mechanistic
understanding can contribute to the debate on the relative merits of
different insecticide strategies extending existing frameworks
{[}12{]}{[}4{]}{[}5{]}{[}1{]}.

Our modelling approach focuses on the change in frequency of a single
resistance gene for each insecticide. It is likely that genes conferring
resistance are present in populations even prior to exposure to novel
insecticides thus leading to the potential for selection {[}14{]}.

\subsection{Methods}\label{methods}

@@

\textbf{Table 1. Parameters inflencing the development of insecticide
resistance}

\begin{longtable}[]{@{}ll@{}}
\toprule
\begin{minipage}[b]{0.28\columnwidth}\raggedright\strut
Parameter\strut
\end{minipage} & \begin{minipage}[b]{0.66\columnwidth}\raggedright\strut
Description\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
1. Effectiveness\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
proportion of susceptible (SS) insects killed by exposure to
insecticide\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
2. Exposure\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
proportion of insects exposed to insecticide\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
3. Resistance restoration\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
ability of resistance (RR) to restore fitness of insects exposed to
insecticide\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
4. Dominance of restoration\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
sets fitness of heterozygous (SR) insects between that of SS \& RR in
presence of insecticide\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
5. Frequency\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
frequency of resistance alleles within the population\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
6. Cost of resistance\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
fitness of resistant (RR) insects in absence of insecticide\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
7. Dominance of cost\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright\strut
sets fitness of heterozygous (SR) insects between that of SS \& RR in
absence of insecticide\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

The calculation of fitness for each genotype when considering just a
single insecticide is illustrated in Figure 1. In the model itself this
is repeated for both insecticides and the fitnesses multiplied. Firstly
the exposure input determines the proportion of the population in the
left and right panels (exposed and not exposed). For those that are
exposed (left panel) insecticide effectiveness sets the fitness for SS
and resistance restoration `restores' a portion of the fitness for RR. A
value of 1 for resistance restoration would bring the fitness of RR back
to 1 (equivalent to the unexposed SS), a value of 0 for resistance
restoration would lead to the RR having the same fitness as SS in the
presence of the insecticide and effectively there would be no
resistance. We use `Resistance restoration' to represent how much the RR
genotype restores fitness in the presence of the insecticide. The
`selection coefficient' of resistance, that may be seen referred to
elsewhere {[}10{]}, can be calculated by multiplying resistance
restoration by insecticide effectiveness. Cost of resistance determines
the fitness of RR in the portion of the population not exposed to the
insecticide.

@@

The model outputs the change in resistance allele frequencies over time
measured in generations. We extract the number of generations at which a
resistance allele frequency of 50\% is reached and term this
`time-to-resistance'. Using a 50\% threshold is consistent with previous
modelling analyses e.g. {[}10{]}{[}15{]}. Preliminary investigations
showed that using other potential thresholds of 25 or 75\% gave very
similar results.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  single insecticide
\item
  two insecticides used in sequence, replacing the first with a second
  once the frequency of resistance alleles reaches a threshold of 50\%
\item
  two insecticides used in a mixture (concurrently)
\end{enumerate}

The model is implemented in R {[}16{]}, the code is hosted on Github
{[}17{]} and we provide an on-line user interface {[}18{]} (Fig 2)
enabling the reader to change inputs and run the model themselves.

\subsection{Results}\label{results-1}

\subsubsection{Single insecticide}\label{single-insecticide}

For single insecticide use, higher values of insecticide effectiveness,
exposure, resistance restoration and its dominance all resulted in
faster resistance spread (Fig 3 A-D) and thus fewer generations to reach
a resistance threshold of 0.5. This makes intuitive sense as all
increase the impact of the insecticide and thus increase selection for
resistance. Similarly, higher values of starting resistance frequency
also resulted in shorter times to resistance thresholds (Fig 4A) through
the even simpler mechanism that less change is required to reach the
threshold.

@@

\subsubsection{Two insecticides}\label{two-insecticides}

When two insecticides are used we can compare the relative performance
of a mixture with sequential use. Each panel in Figures 5-9 compares
mixture and sequence for a single combination of inputs. To help
understand the mechanisms influencing this relative performance we start
with a single base scenario and demonstrate changing inputs
individually. In this base scenario we set effectiveness, exposure,
resistance restoration, and dominance at 0.5, the starting frequencies
of resistance to 0.01 and cost of resistance to 0. Resistance arises
slower for sequential than mixture for this base scenario (Fig 5A).
Resistance to both insecticides in the mixture follow the same path and
reach the threshold at the same time (as would be expected given that
they have identical input parameters). When used in sequence the curve
(dashed) for each insecticide individually is steeper than in the
mixture but, because they happen one after the other, it takes longer
for both to reach the resistance threshold (Fig 5A). Thus the
time-to-resistance for the mixture divided by the sequence is less than
1 at 0.8.

\paragraph{Insecticide exposure and
effectiveness}\label{insecticide-exposure-and-effectiveness}

@@

The results is that a more effective insecticide increases
time-to-resistance when used in a mixture (compare the solid lines in
Fig 5A \& B). This is opposite to what happens when used in sequence
(compare the red dashed line in Fig 5A \& B) and in isolation (Fig 2A),
when more effective insecticides shortened times-to-resistance.

@@

\paragraph{Resistance restoration and
dominance}\label{resistance-restoration-and-dominance}

Increasing resistance restoration or its dominance decreases
time-to-resistance for both sequences and mixtures (Fig 7) in the same
way that it did for sole use (Fig 3). The result is that increasing
either or both of dominance and resistance restoration (Figs 7B,C,D)
doesn't change the relative time-to-resistance for mixtures and
sequences from that in the base scenario (Fig 7A) where it takes
resistance longer to develop for the sequence. So, unlike effectiveness
and exposure, the levels of resistance restoration and dominance do not
effect whether a mixture or a sequence are likely to be favoured.

\paragraph{Starting frequencies of
resistance}\label{starting-frequencies-of-resistance}

Changing the starting frequency of resistance had a similar effect on
time to resistance for both sequences and mixtures and thus had little
effect on their relative performance. For example taking the base
scenario and reducing the starting frequency of one resistance allele
did not change from sequence being favoured (compare Fig 8C to 8A).
Similarly taking a scenario in which time-to-resistance is longer for a
mixture (Fig 8B) and decreasing the starting frequency of that
resistance allele did not change the fact that the mixture was favoured
(Fig 8D).

\paragraph{Costs of resistance}\label{costs-of-resistance}

Increasing the cost of resistance to one insecticide in a mixture leads
resistance to that insecticide to evolve more slowly (compare the solid
red line to the solid blue line in Figure 9C, the red line is for the
resistance with the higher cost). The resistance with the higher cost
also increases more slowly in the mixture relative to when it is used in
sequence (compare the solid red line to the dotted red line in Fig 9C).
The greater effect of cost on mixtures means the advantage of sequences
over mixtures in Fig 9A (and Fig 5A) is removed in Fig 9C. With costs
set to 0 time-to-resistance for mixture divided by sequence is 0.8 in
Fig 9A and 1 in Fig 9C. At higher insecticide effectiveness the benefit
of mixture over sequence is improved as seen by comparing Fig 9D to 9B.

@@

\paragraph{Linkage disequilibrium}\label{linkage-disequilibrium}

It has previously been suggested {[}10{]}{[}1{]} that using an
insecticide to which there was existing resistance in a mixture with a
new insecticide with little resistance could lead to more rapid
evolution of resistance to the latter through a process termed linkage
disequilibrium. Figure 8 C\&D show how our model predicts that linkage
disequilibrium does not lead to more rapid evolution of resistance to
the newer insecticide. In Figures 8 C\&D the newer insecticide with
lower starting resistance (in red) leads to faster resistance when used
alone (red dotted) than when used in a mixture (red solid).

\subsection{Discussion}\label{discussion}

@@

These mechanisms, explaining the response of resistance to single
insecticide use under different input values, are summarised in Table 2.
Referring to figure 1 can help explain these. Exposure and effectiveness
act by decreasing the fitness of all genotypes, dominance of restoration
acts by increasing fitness of SR and resistance-restoration acts by
increasing fitness of RR \& SR. Cost acts by decreasing the fitness of
RR and SR for those not exposed.

\paragraph{Two insecticides}\label{two-insecticides-1}

Using two insecticides either in sequence or in mixtures differs from
the use of a single insecticide and, where it differs can again help us
to understand the dynamics of resistance evolution. Our base scenario
(Fig 5A) shows curves for resistance to both insecticides in the mixture
that are shallower than those for sequential use. The shallower curves
for the mixture can be explained by each insecticide killing individuals
that are resistant to the other insecticide. Thus each insecticide
reduces the selection pressure for the increase in resistance to the
other and could be said to `protect' the other. However this protection
is not necessarily enough to ensure that resistance arises more slowly
in a mixture relative to a sequential strategy. Resistance can evolve
more slowly for both insecticides in sequential use, despite being
faster for each, because they occur one after the other. In Figure 5A it
takes 60 generations for resistance to reach the threshold for each
insecticide in the sequence, and 90 when they are applied together in
the mixture, so the sequence is still favoured because 2x60 is greater
than 90. Incraesing the effectiveness of on insecticide to Fig 5B
decreases the added times for the sequence to 90 and increases those for
the mixture to 115 thus changing the order and favouring the mixture.

The mechanism for this change can be sen by comparing the shape of the
curves in Figs 5A \& B. In the mixture, resistance to insecticide 1
(with the increased effectiveness) rises at a similar speed in both
figures. However, the increased effectiveness of insecticide 1 causes
resistance to insecticide 2 to increase more slowly initially in the
mixture. Once insecticide 1 reaches resistance of around 50\% at around
50 generations, the curve for insecticide 2 becomes steeper. Once the
first insecticide has become ineffective due to resistance it stops
killing individuals that are resistant to the second insecticide, loses
its protective effect and thus stops slowing the rise in resistance to
the second insecticide. A similar effect is visible when the
effectiveness of both insecticides are increased (Fig 6C). The identical
resistance curves for each insecticide in the mixture are shallower than
at the lower effectiveness (Fig 5A) because more individuals resistant
to each insecticide are being killed by the other insecticide. In this
case the `protection' given to both insecticides by the other declines
at the same rate so we don't see the change in slope seen in Fig 5B.

Fig 5B also demonstrates that when two insecticides have a different
effectiveness in a mixture, other parameters being equal, resistance
will be expected to increase faster to the more effective insecticide.
The more effective insecticide prompts both a) higher selection pressure
for it's own resistance and b) greater `protection' reducing the rise in
resistance to the other insecticide.

\paragraph{Difference between exposure and
effectiveness}\label{difference-between-exposure-and-effectiveness}

It seems initially counter-intuitive that whereas increasing
effectiveness slows the development of resistance for mixtures,
increasing exposure speeds it up. Increasing the exposure to both
insecticides results in a decrease in time-to-resistance (from 90
generations in Fig 5A to 40 generations in Fig 5C) where increasing the
effectiveness of both insecticides by the same amount results in a
slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts
with the identical effects that exposure and effectiveness have on a
single insecticide (Figs 3 A,B) and on insecticides in sequence (the red
dashed lines are the same in Figs 5B \& 5C). This difference, that
exposure reduces times-to-resistance for a mixture and effectiveness
increases it, was unexpected for us.

\paragraph{Costs of resistance}\label{costs-of-resistance-1}
